% !TEX root = 00_dmlr_beyond_scale.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  Beyond Scale: The Diversity Coefficient as a Data Quality Metric
%  DMLR (Data-centric Machine Learning Research) submission
%
%  Organized structure:
%    00_dmlr_beyond_scale.tex  — root file (this file)
%    01_introduction.tex       — Introduction
%    02_method.tex             — Method: Diversity Coefficient
%    03_experiments.tex        — Experiments & Results
%    04_related_work.tex       — Related Work
%    05_discussion.tex         — Discussion
%    98_acknowledgments.tex    — Acknowledgments
%    99_appendix.tex           — Appendix (all supplementary sections)
%
%  Style:   dmlr2e.sty  (from https://github.com/JmlrOrg/dmlr-style-file)
%  Bib:     dmlr_bib_refs.bib
%  Figures: plots/
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[twoside,11pt]{article}
\usepackage{dmlr2e}

% Optional math commands from https://github.com/goodfeli/dlbook_notation
\input{math_commands.tex}

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{url}

% Required by dmlr2e.sty — fill in before camera-ready submission
\def\openreview{\url{https://openreview.net/forum?id=XXXX}}

% Journal heading metadata — fill in before camera-ready submission
% \dmlrheading{Volume}{Year}{pages}{submitted}{published}{id}{authors}

% Running headers
\ShortHeadings{Beyond Scale: The Diversity Coefficient}{Miranda, Lee, Sundar, Casasola, Schaeffer, Obadd \& Koyejo}

\firstpageno{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\title{Beyond Scale: The Diversity Coefficient as a Data Quality Metric for Variability in Natural Language Data}

\author{%
  \name Brando Miranda\thanks{Correspondence to: \texttt{brando9@stanford.edu} and \texttt{sanmi@stanford.edu}}
  \AND
  \name Alycia Lee \AND
  \name Sudharsan Sundar \AND
  \name Allison Casasola \AND
  \name Rylan Schaeffer \AND
  \name Elyas Obadd \AND
  \name Sanmi Koyejo \\
  \addr Department of Computer Science, Stanford University, Stanford, CA 94305, USA \\
  \texttt{\{brando9, alylee15, sjsundar, allyc, rschaef, eobbad, sanmi\}@stanford.edu}
}

\editor{Handling Editor: TBD}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Current trends in pre-training Large Language Models (LLMs) primarily focus on the scaling of model and dataset size.
While the \textit{quality} of pre-training data is considered an important factor for training powerful LLMs, it remains a nebulous concept that has not been rigorously characterized.
To this end, we propose a formalization of one key aspect of data quality -- measuring the \textit{variability} of natural language data -- specifically via a measure we call the diversity coefficient.
Our empirical analysis shows that the proposed diversity coefficient aligns with the intuitive properties of diversity and variability,
e.g., it increases as the number of latent concepts increases.
Then, we measure the diversity coefficient of publicly available pre-training datasets and demonstrate that their formal diversity is high compared to theoretical lower and upper bounds.
Finally, we conduct a comprehensive set of controlled \textit{interventional} experiments with GPT-2 and LLaMAv2 that demonstrate the diversity coefficient of pre-training data characterizes useful aspects of downstream model evaluation performance---totaling 44 models of various sizes (51M to 7B parameters).
We conclude that our formal notion of diversity is an important aspect of data quality that captures variability and causally leads to improved evaluation performance.
\end{abstract}

\begin{keywords}
data quality, data diversity, diversity coefficient, Task2Vec, large language models, pre-training data, natural language processing
\end{keywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main sections
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{01_introduction}

\input{02_method}

\input{03_experiments}

\input{04_related_work}

\input{05_discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Broader Impact Statement (required for DMLR)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\impact{
This work proposes the diversity coefficient as a quantitative metric for data quality in LLM pre-training.
By making data diversity formally measurable, we enable researchers and practitioners to design more principled and transparent data curation pipelines.
A potential concern is that diversity metrics could be gamed or misused to justify data collection practices that raise privacy or copyright issues; we encourage responsible use alongside appropriate data governance.
All datasets used in this study are publicly available.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{98_acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliography{dmlr_bib_refs}
\bibliographystyle{plainnat}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\input{99_appendix}

\end{document}
